Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-16 paper/manuscript.tex
Updates from Overleaf

Free energy information should generally be saved more frequently than coordinate data, approximately at the rate that uncorrelated samples are produced.  
The on-disk size of the data for free energy estimation is often significantly smaller than full atomic coordinates, so the information should be collected frequently. 
However, the information should not be collected \textit{every} time step, as most free energy techniques are operated at equilibrium, and need equilibrated \textit{and decorrelated} samples for an unbiased estimate.
A sample collected every time step will likely result in most samples being discarded due to decorrelation routines in the analysis. However, if it is computationally cheap and disk space is plentiful, do save often. One may safely assume that the correlation time is greater than 100-200 fs even for relatively simple systems such as small molecules in solvent, so saving no more frequently than every 50-100 steps is recommended. 
How decorrelation impacts calculations, and how to compute it is discussed in other sections.  [section ref]
%
\section{Data analysis}
\label{sec:data_analysis}
Once data has been collected from alchemical intermediates, it must be analyzed to produce an estimate of the free energy change (and its associated statistical uncertainty) for each leg of the thermodynamic cycle.
While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
%
\subsection{Detecting the boundary between equilibrated and production regions}
\label{sec:automatic-equilibration-detection}
Much of the infrastructure for analyzing alchemical free energy calculations relies on the concept of asymptotically unbiased estimators, which produce unbiased estimates of the free energy when fed very long simulations~\cite{shirts2005comparison}.
In reality, free energy calculations are often initiated from highly atypical initial conditions (such as a protein-ligand geometry obtained from docking and subjected to a heuristic solvent placement scheme), and simulations are of a finite length dictated by available computational resources and computing demands.
As a result, these estimators can produce significantly biased estimates if fed the entirety of simulation data generated without further processing~\cite{chodera2016simple}.
%
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{figures/fig11_equib_detection/Figure.pdf}
    \caption{{\bf Automatic partitioning into equilibration and production regions.}
    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient before relaxing to the true average potential energy\textbf{B} A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue). Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}) \textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
    \emph{Figure adapted from~\cite{chodera2016simple}.}
    }
    \label{fig:automatic-equilibration-detection}
\end{figure}

To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step.

A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Figure~\ref{fig:automatic-equilibration-detection}).
Suppose we have a simulation of length $T$ consisting of correlated data.
Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
\begin{eqnarray}
N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
\end{eqnarray}
where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
Conveniently, this procedure also produces the information necessary to decorrelate the simulation data for estimating the free energy differences, a requisite next step in analysis.
This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.

@@

\paragraph{Computing the timeseries for equilibration detection}
Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
For simple independent simulations that sample $x_t \sim \pi(x ; \lambda)$, this is given by the reduced potential
\begin{eqnarray}
a_t &\equiv& - \ln \pi(x_t; \lambda) + c = u(x_t; \lambda) .
\end{eqnarray}

@@

\paragraph{Cautions in automating equilibration detection}
For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of conformational space, which may itself lead to biased estimates. 
This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate.
%
\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the free energy difference and its statistical uncertainty.
To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
\begin{eqnarray}
g &\equiv& 1 + 2 \tau_\mathrm{eq} \label{eq:statistical-inefficiency-definition}
\end{eqnarray}
where $\tau_\mathrm{eq}$ is the integrated autocorrelation time, formally defined as
\begin{eqnarray}
\tau_\mathrm{eq} &\equiv& \sum_{t=1}^{T-1} \left(1 - \frac{t}{T}\right) C_t \label{eq:integrated-autocorrelation-time-definition} , 
\end{eqnarray}
with the discrete-time normalized fluctuation autocorrelation function $C_t$ defined as
\begin{eqnarray}
C_t &\equiv& \frac{\expect{a_n a_{n+t}} - \expect{a_n}^2}{\expect{a_n^2} - \expect{a_n}^2} . \label{equation:autocorrelation-definition}
\end{eqnarray}
The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).

>>>>

Free energy information should generally be saved more frequently than coordinate data, approximately at the rate that uncorrelated samples are produced.  
The on-disk size of the data for free energy estimation is often significantly smaller than full atomic coordinates, so the information should be collected frequently. 
However, the information should not be collected \textit{every} time step, as most free energy techniques are operated at equilibrium, and need equilibrated \textit{and decorrelated} samples for an unbiased estimate.
A sample collected every time step will likely result in most samples being discarded due to decorrelation routines in the analysis. However, if it is computationally cheap and disk space is plentiful, do save often. One may safely assume that the correlation time is greater than 100-200 fs even for relatively simple systems such as small molecules in solvent, so saving no more frequently than every 50-100 steps is recommended. 
How decorrelation impacts calculations, and how to compute it is discussed in other section~\ref{sec:decorrelating-samples}. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  Data Analysis               %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data analysis}
\label{sec:data_analysis}
Once data has been collected from alchemical intermediates, it must be analyzed to produce an estimate of the free energy change (and its associated statistical uncertainty) for each leg of the thermodynamic cycle.
While a number of different estimators are available that will give consistent results under optimal circumstances, some approaches are recommended over others due to their robustness and ability to provide information on poor convergence.
\subsection{Detecting the boundary between equilibrated and production regions}
\label{sec:automatic-equilibration-detection}
Much of the infrastructure for analyzing alchemical free energy calculations relies on the concept of asymptotically unbiased estimators, which produce unbiased estimates of the free energy when fed very long simulations~\cite{shirts2005comparison}.
In reality, free energy calculations are often initiated from highly atypical initial conditions (such as a protein-ligand geometry obtained from docking and subjected to a heuristic solvent placement scheme), and simulations are of a finite length dictated by available computational resources and computing demands.
As a result, these estimators can produce significantly biased estimates if fed the entirety of simulation data generated without further processing~\cite{chodera2016simple}.
    \centering
    \includegraphics[width=0.95\linewidth]{figures/fig11_equib_detection/Figure.pdf}
    \caption{{\bf Automatic partitioning into equilibration and production regions.}
    \textbf{A} The average (black line) standard deviation (shaded region) of the reduced potential $u^*$ over many independent replicate simulations started from the same initial conditions show a significant initial transient change before relaxing to the true average potential energy \textbf{B}. A cumulative average (red) of the entire simulation data demonstrates simulation bias not seen when initial simulation data is omitted (blue). Using an automated approach to detect equilibration of the boundary $t_0$ using statistical inefficiency $g$ (\textbf{C})  for an effective simulation interval (\textbf{D}). \textbf{E} The optimal equilibration boundary $t_0$ is selected to maximize the number of uncorrelated samples.
    \emph{Figure adapted from~\cite{chodera2016simple}.}
    \label{fig:automatic-equilibration-detection}
To minimize this effect, an initial portion of the simulation is often discarded to \emph{equilibration}~\cite{braun2019best}, with the idea of removing the most heavily biased initial portion of simulation data but retaining the unbiased \emph{production} region that represents a stationary Markov chain process sampling from the desired equilibrium target distribution.
Because the simulation time required for the atypical initial sampler state to relax toward equilibrium is a property of the specific system being simulated and the specific initial conditions selected, it is simplest to collect data for the whole process and use an automated algorithm to select how much data should be discarded to equilibration in a post-processing step.

A simple approach to automatically partitioning simulation data into equilibration and production regions is described in~\cite{chodera2016simple} (illustrated in Figure~\ref{fig:automatic-equilibration-detection}).
Suppose we have a simulation of length $T$ consisting of correlated data.
Here, the goal of the post-processing step is to select the equilibration boundary $t_0 \in [0, T]$ so as to \emph{maximize} the number of effectively uncorrelated samples remaining in the production region $N_{[t_0,T]}$, which is defined as
N_{[t_0,T]} &=& \frac{T - t_0}{g_{[t_0,T]}}
where $g_{[t_0,T]}$ is the \emph{statistical inefficiency} of a timeseries $a_t$, described in more detail below.
Conveniently, this procedure also produces the information necessary to decorrelate the simulation data for estimating the free energy differences, a requisite next step in analysis.
This approach is implemented within the MBAR~\cite{kylebeauchamp2019choderalab} and alchemlyb~\cite{daviddotson2019alchemistry} packages, and is highly recommended for standard practice.

@@

\paragraph{Computing the timeseries for equilibration detection}
Typically, the timeseries of note $a_t$ analyzed in automated equilibration detection is the negative logarithm of the probability density sampled by the MCMC algorithm (up to an irrelevant additive constant).
For simple independent simulations that sample $x_t \sim \pi(x ; \lambda)$, this is given by the reduced potential
a_t &\equiv& - \ln \pi(x_t; \lambda) + c = u(x_t; \lambda) .

@@

\paragraph{Cautions in automating equilibration detection}
For simulations that are simply not long enough to contain a large number of samples from true equilibrium either because they are very short or contain slow processes, this procedure cannot completely remove the bias.
In such cases, this approach simply selects the final portion of the the simulation, which may be contained in a single substate of conformational space, which may itself lead to biased estimates. 
This situation can be detected if the equilibration boundary $t_0$ is a significant fraction of the total simulation length $T$, with a good rule of thumb being that $T \gtrsim 20 t_0$.
If this is not possible, advanced analysis techniques that assume only local equilibrium (rather than global equilibrium) such as the TRAM estimators~\cite{mey2014xtram,wu2016multiensemble,nuske2017markov} may be more appropriate, but are beyond the scope of this paper. 
\subsection{Decorrelating samples for analysis}
\label{sec:decorrelating-samples}
\paragraph{Computing the statistical inefficiency}
Most estimators require an uncorrelated set of samples from the equilibrium distribution to produce (relatively) unbiased estimates of the free energy difference and its statistical uncertainty.
To do this, the production region of the simulation is generally \emph{subsampled} with an interval approximately equal to or greater than the \emph{statistical inefficiency} $g \ge 1$ to produce a set of uncorrelated samples that can be fed to the estimator machinery~\cite{chodera2016simple},
g &\equiv& 1 + 2 \tau_\mathrm{eq} \label{eq:statistical-inefficiency-definition}
where $\tau_\mathrm{eq}$ is the integrated autocorrelation time, formally defined as
\tau_\mathrm{eq} &\equiv& \sum_{t=1}^{T-1} \left(1 - \frac{t}{T}\right) C_t \label{eq:integrated-autocorrelation-time-definition} , 
with the discrete-time normalized fluctuation autocorrelation function $C_t$ defined as
C_t &\equiv& \frac{\expect{a_n a_{n+t}} - \expect{a_n}^2}{\expect{a_n^2} - \expect{a_n}^2} . \label{equation:autocorrelation-definition}
The basic concept is that $\tau_\mathrm{eq}$ corresponds to the single-exponential decay time for the autocorrelation process that generates samples, so the statistical inefficiency $g$ measures the approximate temporal separation between two effectively uncorrelated samples (where two exponential relaxation times are presumed to be sufficient).
